---
title: "Getting Started"
author: "Anders H. Jarmund"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%", 
  fig.asp = 0.7,
  fig.width = 12,
  fig.align = "center",
  cache = FALSE,
  external = FALSE
)
library("ALASCA")
library("data.table")
library("ggplot2")
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

# Introduction to ALASCA

<script src="js/hideOutput.js"></script>
<style>
.showopt {
  background-color: #004c93;
  color: #FFFFFF; 
  width: 150px;
  height: 20px;
  text-align: center;
  vertical-align: middle !important;
  float: left;
  font-family: sans-serif;
  border-radius: 8px;
}

.showopt:hover {
    background-color: #dfe4f2;
    color: #004c93;
}

.showopttext::before{
  content: "Show/Hide Source"
}

pre.plot {
  background-color: white !important;
}
</style>

The [ALASCA package](https://andjar.github.io/ALASCA) is described in the paper [ALASCA: An R package for longitudinal and cross-sectional analysis of multivariate data by ASCA-based methods.](https://www.frontiersin.org/articles/10.3389/fmolb.2022.962431/full). The paper contains several examples of how the package can be used.

This vignette will only show how to quickly get started with the ALASCA package. For more examples, see

* [Regression models](articles/regression.html)
* [Plotting](articles/plot.html)

## Installation

```
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

devtools:install_github(“andjar/ALASCA”, ref = “main”)
```

### Citation
If you have utilized the ALASCA package, please consider citing:

Jarmund AH, Madssen TS and Giskeødegård GF (2022) ALASCA: An R package for longitudinal and cross-sectional analysis of multivariate data by ASCA-based methods. *Front. Mol. Biosci.* 9:962431. doi: 10.3389/fmolb.2022.962431

```
@ARTICLE{10.3389/fmolb.2022.962431,
  AUTHOR={Jarmund, Anders Hagen and Madssen, Torfinn Støve and Giskeødegård, Guro F.},
  TITLE={ALASCA: An R package for longitudinal and cross-sectional analysis of multivariate data by ASCA-based methods},
  JOURNAL={Frontiers in Molecular Biosciences},
  VOLUME={9},
  YEAR={2022},
  URL={https://www.frontiersin.org/articles/10.3389/fmolb.2022.962431},       
  DOI={10.3389/fmolb.2022.962431},      
  ISSN={2296-889X}
}
```

## Creating an ASCA model

### Generating a data set

We will start by creating an artificial data set with 100 participants, 5 time points, and 20 variables. The variables follow four patterns

* Linear increase
* Linear decrease
* A v-shape
* An inverted v-shape

<div class="fold s o">
```{r}
n_time     <- 5
n_id       <- 100
n_variable <- 20

df <- rbindlist(lapply(seq(1,n_id), function(i_id) {
  rbindlist(lapply(seq(1,n_variable), function(i_variable) {
    
    r_intercept <- rnorm(1, sd = 5)
    beta <- 2 + rnorm(1)
    
    temp_data <- data.table(
                  id = paste0("id_", i_id),
                  time = seq(1, n_time) - 1,
                  variable = paste0("variable_", i_variable)
                )
    if ((i_variable %% 4) == 0) {
      temp_data[, value := r_intercept + beta * time]
    } else if ((i_variable %% 4) == 1) {
      temp_data[, value := r_intercept - beta * time]
    } else if ((i_variable %% 4) == 2) {
      temp_data[, value := r_intercept - beta*n_time/2 + beta * abs(time - n_time/2)]
    } else {
      temp_data[, value := r_intercept + beta*n_time/2 - beta * abs(time - n_time/2)]
    }
    
    temp_data[, value := value + rnorm(n_time)]
    temp_data[, value := value * i_variable/2]
    temp_data
  }))
}))
```
</div>

Overall (ignoring the random effects), the four patterns look like this:

```{r}
ggplot(df[variable %in% c("variable_1", "variable_2", "variable_3", "variable_4"),],
       aes(time, value)) +
  geom_smooth() +
  facet_wrap(~variable, scales = "free_y")
```

### Data format

We want time to be a categorical variable:

```{r}
df[, time := paste0("t_", time)]
```

Your data can either be provided in long or wide format. In long format, there is one column with variable names and one column with the variable values. For example:

```{r}
head(df)
```

In wide format, each variable has a separate column:

```{r}
head(dcast(data = df, ... ~ variable))
```

ALASCA supports both formats but defaults to long format. To use wide format, you have to set `wide = TRUE`. 

### Initialize an ALASCA model

In this example, we are only looking at the common time development. For examples involving group differences, see [the vignette on regression models](articles/regression.html).

To assess the time development in this data set, we will use the regression formula `value ~ time + (1|id)`. Here, `value` is the measured variable value, `time` the predictor, and `(1|id)` a random intercept per participant-id. ALASCA will implicitly run the regression for each variable separately.

```{r}
res <- ALASCA(
  df,
  value ~ time + (1|id)
)
```

The ALASCA function will provide output with important information:

* `Guessing effects: 'time'` When effects are not explicitly provided to ALASCA, the package will try to guess the effects you are interested in. See [the vignette on regression models](articles/regression.html) for details.
* `Will use linear mixed models!` ALASCA will use linear mixed models when you provide a random effect in the regression formula (i.e., `(1|id)`)
* `Will use Rfast!` Linear mixed model regression can be performed by one out of two different R packages: the [lme4 package](https://cran.r-project.org/web/packages/lme4/index.html) or the [Rfast package](https://cran.r-project.org/web/packages/Rfast/index.html)
* `The 'time' column is used for stratification` This is only important for model validation. For details, see [the vignette on model validation](articles/validation.html)
* `Converting 'character' columns to factors` We provided time as a character variable and ALASCA converts it to a factor variable. If the levels of your variable matters and they are not in alphabetical order, you may want to convert the variable to a factor by yourself.
* `Scaling data with sdall ...` ALASCA supports various scalings, and `sdall` is the default. For details, see our paper [ALASCA: An R package for longitudinal and cross-sectional analysis of multivariate data by ASCA-based methods.](https://www.frontiersin.org/articles/10.3389/fmolb.2022.962431/full)
* `Calculating LMM coefficients` Simply informs you that the regression is ongoing as this may take some time


To see the resulting model:

```{r}
plot(res, component = c(1,2), type = 'effect')
```

See [the vignette on plotting the model](articles/plot.html) for more visualizations.
