---
title: "Personality and Aging"
author: "Anders H. Jarmund"
date: "`r Sys.Date()`"
bibliography: ALASCA.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Personality and Aging}
  %\VignetteEngine{knitr::rmarkdown}
  %VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%", 
  fig.asp = 0.7,
  fig.width = 12,
  fig.align = "center",
  cache = FALSE,
  external = FALSE
)
library("kableExtra")
```

# Preparing the data

<script src="js/hideOutput.js"></script>
<style>
.showopt {
  background-color: #004c93;
  color: #FFFFFF; 
  width: 150px;
  height: 20px;
  text-align: center;
  vertical-align: middle !important;
  float: left;
  font-family: sans-serif;
  border-radius: 8px;
}

.showopt:hover {
    background-color: #dfe4f2;
    color: #004c93;
}

.showopttext::before{
  content: "Show/Hide Source"
}

pre.plot {
  background-color: white !important;
}
</style>

```{r setup}
library("reshape2")
library("ALASCA")
```

This is a demonstration of single measures data analyzed with `ALASCA()`, with personality data explored with focus on age and gender. You may also be interested in

* Introduction to ALASCA, see [Get Started](ALASCA.html)
* Analysis of **interventional**, see [the vignette on metabolomics](metabolomics.html)
* Analysis of **observational** data, see [the vignette on early preeclampsia](pregnancy.html)

We will use the *Development of the Firstborn Personality Scale* data from the [Open-Source Psychometrics Project](https://openpsychometrics.org/_rawdata/). It is quite large, so we will restrict ourself to participants from Great Britain (to speed up things). Let us import and prepare the data:

<div class="fold s">
```{r import_data}
df <- read.csv(file = "personality.csv", header = TRUE, sep = ";")
df$ID <- 1:nrow(df)
df <- melt(df, id.vars = c("ID","age", "engnat", "gender", "birthpos", "birthn", "submittime", "country", "dateload", "source", "screensize", "introelapse", "testelapse", "endelapse"))

df <- subset(df, country == "GB")

# Notice that time is actually age spans
df$time <- ifelse(df$age < 20, "<20", 
                  ifelse(df$age < 40, "20-40", 
                         ifelse(df$age < 60, "40-60", ">60")))
df$time <- factor(df$time, levels = c("<20", "20-40", "40-60", ">60"))
df$group <- ifelse(df$gender == 1, "Male", ifelse(df$gender == 2, "Female", NA))
df <- subset(df, !is.na(group))
head(df)
```
</div>

Note that the question text itself is kept in another variable for the time being,

```{r import_data_2}
df.q <- read.csv(file = "personality_q.csv", header = TRUE, sep = ";")
colnames(df.q) <- c("code", "question") # to get rid of ï..
head(df.q)
```

# Applying ALASCA

Applying `ALASCA()` is quite straightforward. If we omit the random effect from our regression model, ALASCA will automatically select linear models.

```{r}
model.formula <- value ~ time*group
mod <- ALASCA(df, model.formula, validate = TRUE, participantColumn = "ID", plot.xlabel = "Age")

plot(mod)
plot(mod, component = 2)
```

We can see that women (reference group) has an increasing trend of *something*. Let us see for which questions their score is increasing and decreasing with age:

```{r}
l_PC1 <- subset(getLoadings(mod)$time, PC == 1) # Get PC1 for time effect
l_PC1 <- merge(l_PC1, df.q, by.x = "covars", by.y = "code", all.x = TRUE) # attach question
l_PC1 <- l_PC1[order(l_PC1$loading, decreasing = TRUE),] # Order them after loading
knitr::kable(l_PC1[1:10, c("loading","question")]) %>%
  kable_material(c("striped", "hover"))
```

And in the other end (in this case, older women score lower on these items):

```{r}
l_PC1 <- l_PC1[order(l_PC1$loading, decreasing = FALSE),] # Order them after loading
knitr::kable(l_PC1[1:10, c("loading","question")]) %>%
  kable_material(c("striped", "hover"))
```

Which features distinguish men from the women? From the plots above, it seems that PC1 indicates a stable difference that do not vary much with age; we do as above; men score higher on these items:

```{r}
l_PC1 <- subset(getLoadings(mod)$group, PC == 1) # Get PC1 for time effect
l_PC1 <- merge(l_PC1, df.q, by.x = "covars", by.y = "code", all.x = TRUE) # attach question
l_PC1 <- l_PC1[order(l_PC1$loading, decreasing = TRUE),] # Order them after loading
knitr::kable(l_PC1[1:10, c("covars","loading","question")]) %>%
  kable_material(c("striped", "hover"))
```

and lower on these

```{r}
l_PC1 <- l_PC1[order(l_PC1$loading, decreasing = FALSE),] # Order them after loading
knitr::kable(l_PC1[1:10, c("covars","loading","question")]) %>%
  kable_material(c("striped", "hover"))
```

We can have a quick look at what out underlying linear model will estimate for the two genders for AGR4;

```{r}
plotPred(mod, variable = "AGR4")
```