---
title: "Plotting"
author: "Anders H. Jarmund"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%", 
  fig.asp = 0.7,
  fig.width = 12,
  fig.align = "center",
  cache = FALSE,
  external = FALSE
)
library("ALASCA")
library("data.table")
library("ggplot2")

n_time     <- 5
n_id       <- 100
n_variable <- 20

df <- rbindlist(lapply(seq(1,n_id), function(i_id) {
  rbindlist(lapply(seq(1,n_variable), function(i_variable) {
    
    r_intercept <- rnorm(1, sd = 5)
    i_group <- i_id %% 2
    if (i_group == 1) {
      beta <- 2 + rnorm(1)
    } else {
      beta <- 3 + rnorm(1)
    }
    
    temp_data <- data.table(
                  id = paste0("id_", i_id),
                  group = paste0("group_", i_group),
                  time = seq(1, n_time) - 1,
                  variable = paste0("variable_", i_variable)
                )
    if ((i_variable %% 4) == 0) {
      temp_data[, value := r_intercept + beta * time]
    } else if ((i_variable %% 4) == 1) {
      temp_data[, value := r_intercept - beta * time]
    } else if ((i_variable %% 4) == 2) {
      temp_data[, value := r_intercept + beta * abs(time - n_time/2)]
    } else {
      temp_data[, value := r_intercept - beta * abs(time - n_time/2)]
    }
    
    temp_data[, value := value + rnorm(n_time)]
    temp_data[, value := value * i_variable/2]
    temp_data
  }))
}))

df[, time := paste0("t_", time)]

res <- ALASCA(
  df,
  value ~ time*group + (1|id)
)

res_2 <- ALASCA(
  df,
  value ~ time*group + (1|id),
  separate_effects = TRUE
)
```

# Visualizations in ALASCA

<script src="js/hideOutput.js"></script>
<style>
.showopt {
  background-color: #004c93;
  color: #FFFFFF; 
  width: 150px;
  height: 20px;
  text-align: center;
  vertical-align: middle !important;
  float: left;
  font-family: sans-serif;
  border-radius: 8px;
}

.showopt:hover {
    background-color: #dfe4f2;
    color: #004c93;
}

.showopttext::before{
  content: "Show/Hide Source"
}

pre.plot {
  background-color: white !important;
}
</style>

We have aimed to implement a range of plotting options for ALASCA. In general, the plotting function is

```
plot(<object>, component = <component>, effect = <effect>, type = <type>, ...)
```

where `<object>` is your ALASCA model, `<component>` and `<effect>` the components and effects of interest, and `<type>` the plot type.

ALASCA can save the plots directly if you specify `save = TRUE` when you initialize the ALASCA model. Otherwise, the plots are returned as ordinary `ggplot2` objects and can be saved with `ggsave()`.

Note that you can specify properties of the plot during plotting, e.g. specifying `palette`:

```{r}
plot(res, component = 1, effect = 1, type = 'effect', palette = c("red", "blue"))
```


## The effect plot

The effect plot is the default plot showing both *scores* and *loadings* for the selected components/effects. You can select one effect and one component:

```{r}
plot(res, component = 1, effect = 1, type = 'effect')
```

Or two components:

```{r}
plot(res, component = c(1, 2), effect = 1, type = 'effect')
```

Or two effects (note: this requires a model with multiple effects):

```{r}
plot(res_2, component = 1, effect = c(1, 2), type = 'effect')
```

Or even a combination of multiple components and effects:

```{r}
plot(res_2, component = c(1,2), effect = c(1, 2), type = 'effect')
```

### Plotting scores or loadings

The effect plot consists of two parts: scores and loadings. You can get those components directly:

```{r}
plot(res, component = 1, effect = 1, type = 'score')
```

And similarly for loadings:

```{r}
plot(res, component = 1, effect = 1, type = 'loading')
```

## Grayscale

Some journals require plots in grayscale (or require fees for colors). You can therefore ask for grayscale figures in ALASCA:

```{r}
res_3 <- ALASCA(
  df,
  value ~ time*group + (1|id),
  plot.grayscale = TRUE
)
plot(res_3, component = 1, effect = 1, type = 'effect')
```

## Adapting the theme

You can specify theme etc. of the plots when you initialize the ALASCA model:

```{r}
res_4 <- ALASCA(
  df,
  value ~ time*group + (1|id),
  plot.my_theme = theme_linedraw(),
  plot.palette = c("red", "blue")
)
plot(res_4, component = 1, effect = 1, type = 'effect')
```

The color palette is by default made from the viridis palette, ending at 0.8: `scales::viridis_pal(end = 0.8)(2)`. You can of course use the colors you want;

```{r, results='hide'}
res_5 <- ALASCA(
  df,
  value ~ time*group + (1|id),
  plot.my_theme = theme_light(),
  validate = TRUE,
  n_validation_runs = 10,
  plot.palette = scales::brewer_pal(type = "qual")(2)
)
plot(res_5, component = 1, effect = 1, type = 'effect')
```

To remove the ribbon, you can set `plot.ribbon = FALSE`:

```{r}
res_5 <- ALASCA(
  df,
  value ~ time*group + (1|id),
  plot.my_theme = theme_light(),
  validate = TRUE,
  n_validation_runs = 10,
  plot.palette = scales::brewer_pal(type = "qual")(2),
  plot.ribbon = FALSE
)
plot(res_5, component = 1, effect = 1, type = 'effect')
```

You can also change these properties directly in the `plot()` function:

```{r}
plot(res_5,
     component = 1,
     effect = 1,
     type = 'effect',
     ribbon = TRUE,
     palette = scales::hue_pal()(2),
     my_theme = theme_classic())
```

or (`x_angle` controls the rotation of the x axis labels, and `x_h_just` is used to center them horizontally)

```{r}
plot(res_5,
     component = 1,
     effect = 1,
     type = 'effect',
     grayscale = TRUE,
     x_angle = 0, x_h_just= 0.5
)
```

Some even prefer to have loadings along the x axis:

```{r}
plot(res_5,
     component = 1,
     effect = 1,
     type = 'effect',
     x_angle = 90, x_h_just = 1, x_v_just = 0.5,
     flip_axis= FALSE
)
```

## Screeplot

Note: The plotting function remembers previous choices, so if you plotted component 1 in the previous plot, you have to specify which components to plot:

```{r}
plot(res, component = seq(1,10), effect = 1, type = 'scree')
```

## Predictions / Marginal means

By default, only some variables are selected based on their loadings:

```{r}
plot(res, component = 1, effect = 1, type = 'prediction')
```

You can ask for specific variables:

```{r}
plot(res, component = 1, effect = 1, type = 'prediction', variable = c("variable_8", "variable_13", "variable_15"))
```

Or adjust the number of retrieved variables (note that we have to set `variable = c()` for the plot function to forget the previous setting):

```{r}
plot(res, component = 1, effect = 1, type = 'prediction', n_limit = 4, variable = c())
```

## Histogram

Visualize the distribution of the bootstrap runs:

```{r, results='hide'}
res_6 <- ALASCA(
  df,
  value ~ time*group + (1|id),
  validate = TRUE,
  n_validation_runs = 100
)
plot(res_6, component = 1, effect = 1, type = 'histogram', n_bins = 30)
```